---
title: "Lab 2b"
subtitle: "Data Visualization"
author: "Statistical programming with R"
params:
  answers: false
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: paged
    number_sections: false
    theme: paper
    highlight: tango
    code_folding: hide 
editor_options: 
  markdown: 
    wrap: 72
---

```{r include = FALSE}
knitr::opts_chunk$set(eval = params$answers)
```

------------------------------------------------------------------------

For this practical, you will need to load the packages `ggplot2`,
`dplyr`, and `patchwork`. If displaying multiple plot in one figure, use
the R chunk options `fig.width` (default = 7), `fig.height` (default =
5) and/or `fig.asp` (height:width ratio) to display the plots in a nice
format. To avoid messages by `ggplot()` being printed to the html, set
the global chunk option to `message = FALSE`

```{r message=FALSE}
library(ggplot2)
library(dplyr)
library(patchwork)
knitr::opts_chunk$set(message = FALSE)
```

For a quick reference of `ggplot2` functions, consult

-   [ggplot2
    Cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-visualization.pdf)

-   [ggplot2
    Reference](https://ggplot2.tidyverse.org/reference/index.html)

------------------------------------------------------------------------

# The diamonds data

The dataset for this practical is the `diamonds` dataset from the
package `ggplot2`.

a.  Check the help page of this `tibble` for its content, and display
    the first 10 rows of the data in your HTML.

```{r}
diamonds
```

The `diamonds` data set has more than 50.000 rows, which is a lot. To
simplify matters, we will work with a random sample of 500 cases.

b.  Set the seed to 100, and randomly sample 500 cases from the dataset
    `diamonds` using the `slice_sample()` function. Save the resulting
    data set as `my_diamonds`.

```{r}
set.seed(100)
my_diamonds <- slice_sample(diamonds, n = 500)
```

# One variable

In the following exercises we display plots numeric and of
categorical/discrete variables. Check the `geoms` overview in the
lecture sheets for the different options.

a.  In a 1 x 3 plot array, plot the distribution of variable `carat`
    using the three different geoms. Add a title with the name of the
    appropriate geom to each plot. The chunk option `fig.asp = .35`
    displays the plots in a nice format.

```{r fig.asp = 0.35}
p <- ggplot(my_diamonds, aes(x = carat))

p + geom_histogram(bins = ) + 
    ggtitle("histogram") +

p + geom_density() + 
    ggtitle("density plot") +
  
p + geom_boxplot() + 
    ggtitle("box plot") 

```

Discrete variables are variables with countable values (i.e. categorical
variables or variables with a small number of integer values). The most
common option for such variables is the bar plot

a.  Display the bar plots of the factors `cut`, `color` and `clarity` in
    a 1 x 3 plot array. Make sure that the labels of the categories are
    readable (i.e., not overlapping).

```{r fig.asp = 0.35}
  ggplot(my_diamonds, aes(y = cut)) + 
    geom_bar() +
  
  ggplot(my_diamonds, aes(y = color)) + 
    geom_bar() +
  
  ggplot(my_diamonds, aes(y = clarity)) + 
    geom_bar() 
```

# Two variables

The combinations between numeric and categorical/discrete variables
requires different geoms. Check the cheat sheet again for the appropriate
options.

a.  In the first row of one figure, display a box and violin plot of the
    variables `price` and `cut`, an in the second row a density plot.
    Use the `/` symbol to start a new row.

```{r}
(ggplot(my_diamonds, aes(price, cut)) + 
    geom_boxplot() +
  
ggplot(my_diamonds, aes(price, cut)) + 
    geom_violin() ) /
  
ggplot(my_diamonds, aes(price, fill = cut)) + 
    geom_density(alpha = .3)  
```

The variables `color` and `clarity` are both categorical variables, and
their distribution is generally displayed by bar plots.

b.  Display a stacked, a dodged, and a filled bar plot of these
    variables in a 1 x 3 plot array. Use the `plot_layout()` function to
    display the legend and axes labels only once.

```{r fig.asp = .35}
ggplot(my_diamonds, aes(color, fill = clarity)) + 
    geom_bar() +
  
ggplot(my_diamonds, aes(color, fill = clarity)) + 
    geom_bar(position = "dodge") +
  
ggplot(my_diamonds, aes(color, fill = clarity)) + 
    geom_bar(position = "fill") +
  
  plot_layout(guides = "collect", axes = "collect")
```

c.  Now display the joint distribution of these two variables with
    `geom_count()`.

```{r}
ggplot(my_diamonds, aes(color, clarity)) + 
    geom_count()
```


# Three variables

The most common plot is that with two numeric variables and one categorical variable, with the levels of the categorical variable displayed with different shapes or colors.

a.  Display two scatter plots of `price` and `carat` in a 1 x 2 plot
    array, once with different sizes for `price`, and once with
    different shapes for `cut`. Make the points transparent by setting
    `alpha` to an appropriate value.

```{r fig.asp = 0.4}
ggplot(my_diamonds, aes(price, carat, size = price)) +
  geom_point(alpha = .1) +

ggplot(my_diamonds, aes(price, carat, col = cut)) +
  geom_point(alpha = .5) 


```

b. The plot with different colors for `cut` is not very clear.
Reproduce the plot by adding a `facet_wrap()` layer, and add `price` for
the size of the points.

```{r}
ggplot(my_diamonds, aes(price, carat, col = cut)) +
  geom_point(aes(size = price), alpha = .5) +
  facet_wrap(vars(cut))
```

d. Display the previous plot with a different theme (choose one that you think looks nice).

```{r}
ggplot(my_diamonds, aes(price, carat, col = cut)) +
  geom_point(aes(size = price), alpha = .5) +
  facet_wrap(vars(cut)) +
  theme_classic()
```


# Advanced

a. Display a graphical representation of the variables `price`, `carat`, `depth`, `cut` and `color`. 

```{r}
ggplot(my_diamonds, aes(price, carat, col = depth)) +
  geom_point(aes(size = color), alpha = .5) +
  facet_wrap(vars(cut)) +
  theme_minimal()
```


------------------------------------------------------------------------

End of lab
