---
title: "Lab 1c exercises"
subtitle: "R data objects"
author: "Statistical Programming with R"
params: 
   answers: true
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: paged
    number_sections: false
    theme: paper
    highlight: tango
    code_folding: hide 
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
knitr::opts_chunk$set(eval = params$answers, comment="")
```

## Opening RStudio Projects

There are several ways to **open an RStudio Project**:

1. Using File -> Open Project in the File menu
2. Selecting a project from the list of most recently opened projects by using File -> Recent Projects
3. Go to the Project folder and double click on the file with the extension .RProject.

Open RStudio and open the RStudio Project you made in the previous lab by choosing one of the three options described above. 

## Assigning values

In the R language the assignment operator `<-` is used to assign values to objects. In the code below you see examples of assigning values. 
```
weight_kg <- 70
height_m <- 1.80
bmi <- weight_kg/(height_m^2)
weight_kg <- 62
```
What is the value of `bmi` after running the four lines of code? 

Remember, to actually RUN the code in your console, you have two options:

* Click on the green triangle in the code chunk.
* If you want to run the code line by line, highlight the code and hit Control-Enter on a PC or Command-Return on a Mac.

```{r}
weight_kg <- 70
height_m <- 1.80
bmi <- weight_kg/(height_m^2)
weight_kg <- 62

# Solution:
# The value of bmi is 21.6 because it is the result of 70/(1.8^2). Changing the 
# weight_kg afterwards did not affect the bmi object.
```


# Vectors

a. Make the vector `v1` containing the numbers 1 to 10. Display the result.

```{r}
v1 <- 1:10
v1
```

b. Make the vector `v2` containing five times the vector c("cat", "dog"). Display the result.

```{r}
v2 <- rep(c("cat", "dog"), 5)
v2
```

c. Make the vector `v3` with `TRUE` and `FALSE` each repeated five times. Display the result.

```{r}
v3 <- rep(c(T, F), each = 5)
v3
```

# Matrices

a. Collect the vectors `v1`, `v2` and `v3` in the matrix `V`, and display the result.

```{r}
V <- cbind(v1, v2, v3)
V
```

b. Display a matrix with five rows and 2 columns with the values 1 to 10.

```{r}
matrix(1:10, 5, 2)
```

# Data frames

a. Make a data frame `df` with the vectors `v1`, `v2` and `v3`, and display it.

```{r}
df <- data.frame(v1, v2, v3)
df
```

b. Make the same `df` with the variable `v2` as `factor` (overwrite the previous version), and display.

```{r}
df <- data.frame(v1, v2 = factor(v2), v3)
df
```

c Find out what the function `str()` does (ask for help about this function).

```{r}
?str
```
Now use the function `str()` to have a look at the `df` data frame you created in the previous exercise. What information does it give about the elements of the data frame?

```{r}
str(df)
```


# Lists

a. Make the list `L` containing the elements "vector" is `v1`, "factor" is `v2` as factor, "matrix" is `V`, and "dataframe" is `df`, and display `L`. 

```{r}
L <- list(vector = v1, factor = factor(v2), matrix = V, dataframe = df)
L
```

# Customizing chunk options 

You can customize the way your R code is displayed in the knitted html file. Options can be added locally or globally.

**Local chunk options** 

Add local chunk options to the header of a code chunk `{r <add options here>}` and the option will hold for this chunk only.

Some commonly used local options:

* Show both R code and output. This is the default R code chunk setting.
* Show the results, not the R code. Use `{r echo = FALSE}` in the code chunk header.
* Do not show the R code, do not show the results. Use `{r include = FALSE}`

Some other useful options are:

* `{r message = FALSE}`: this prevents messages generated by code from appearing in your knitted file.
* `{r warning = FALSE}`: this prevents warning messages to appear in your knitted file.

**Global options** 

To set global options that apply to every chunk in your file, call `knitr::opts_chunk$set` in a code chunk, preferably on top of your R Markdown document. The following code sets the following global defaults for each code chunk, but this global default can be overwritten by options specified in individual chunk headers.

```
knitr::opts_chunk$set(
  echo       = FALSE, # whether to show code chunks
  message    = FALSE, # whether to show messages from your code
  warning    = FALSE, # whether to show warnings from your code
)
```

See the [R Markdown Cheat Sheet](https://posit.co/wp-content/uploads/2022/10/rmarkdown-1.pdf) for a complete list of knitr chunk options.


**Question 1**

Use the code chunk below to explore some local chunk options, such as `{r echo = FALSE}` or `{r include = FALSE}` and see what happens when you knit the this R Markdown file. 

```{r }

df <- data.frame(v1, v2, v3)
summary(df)
```

# Advanced

Set a global chunk option at the beginning of this document: the code that was provided in the explanation above and see what the results are when you knit this document. Check what happens to the individual code chunk option you provided in the previous exercise.



