---
title: "Lab 4b"
subtitle: "Statistical inference, bootstrap"
author: "Utrecht Summer School: Statistical Programming with R"
params: 
   answers: true
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
    number_sections: true
    theme: paper
    highlight: tango
    code_folding: hide 
editor_options: 
  markdown: 
    wrap: 72
---

```{r global_opts, include=FALSE}
knitr::opts_chunk$set(eval = params$answers,
                      message = FALSE, 
                      warning = FALSE,
                      comment = "",
                      fig.width = 16/2,
                      fig.height = 9/2)

```

------------------------------------------------------------------------

# Bootstrap

In this exercise we you will work with the bootstrap function presented in the lecture slides.

## Load the necessary packages

Load the packages: `tidyverse`, `ggplot2` and `patchwork`. The package `pacman` helps to manage loading your packages. 


```{r Load_packages, echo=FALSE}
# Ensures the package "pacman" is installed
if (!require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse,
               ggplot2,
               patchwork)
```


##  Create an R script for the bootstrap function using the following steps:

* Open an R script in RStudio with: File -> New File -> R Script
* Copy-paste the bootstrap function from the lecture in the R script file and save it as `my_bootstrap`, 
* Source the `my_bootstrap` file by running the the in the R chunk below. Check if the function appeared in your workspace.

```{r}
source("my_bootstrap.R")
```

## Load the data

Load the mammals sleep data with the `load()` function. The data are in the folder "Lecture and Lab 4b" and are in the R data format (extension .Rdata). Check whether you see the data in the workspace.

```{r}
load("MammalsSleepData.RData")
```

## Inspect the data

Have a quick inspection of the data with `glimpse()` or `str()` and look at the descriptive statistics with `summary()`

```{r}
glimpse(mammals)
# str(mammals)
```

```{r}
summary(mammals)
```


## Create a subset of the data

Create a subset of the data with only the variables: `total_sleep`, `brain_wt`, `body_wt` and `danger`, and remove the missing values (hint: use the pipe operator).

```{r}
subset_mammals <- mammals %>% 
  select(total_sleep, brain_wt, body_wt, danger) %>%
  na.omit()
```

## Create a new variable

Create the variable `brain_weight` which is the ratio of brain weight and body weight. Compute the new variable as follows: `brain_weight = brain_wt / body_wt` and save the new variable to the subset_mammals data.

```{r}
subset_mammals <- subset_mammals %>% 
  dplyr::mutate(brain_body = brain_wt / body_wt)
```


## Fit a linear model 

Fit the linear model `total_sleep ~ brain_weight + danger` of the `subset_mammals` data and save the `lm` object as `fit_mammals`.

```{r}
fit_mammals <- lm(total_sleep ~ brain_body + danger, data = subset_mammals)
```

Display the summary of this linear model and save it. What do you conclude about the results?

```{r}
sum_fit_mammals <- summary(fit_mammals)
sum_fit_mammals
```
Obtain the 95% confidence intervals for the regression coefficients of the linear model and save the results.

```{r}
CI95 <- confint(fit_mammals) 
CI95
```


##  Run the bootstrap 

Run the bootstrap function on the same linear regression model as in the previous question. Save the results as `boot_mammals`. Do not forget to set a random number seed.

```{r}
set.seed(848)
boot_mammals <- bootstrap(total_sleep ~ brain_body + danger, data = subset_mammals) 
```


## Visualize the bootstrap distributions

Make a histogram of the bootstrap distributions for the predictors `brain_body` and `danger` using the package `ggplot2`. 

The package `ggplot2` works best with data frames. The output of the bootstrap function, `boot_mammals` is a list. Convert the list into a data frame first. 

```{r}
# Make dataframe from the list of bootstrap coefficients
df_boot_mammals <- data.frame(boot_mammals)
```

Make a histogram of the bootstrap distribution for the regression coefficient of `brain_body` and for `danger` with `ggplot2`. 
```{r}

df_boot_mammals %>%
  ggplot(aes(brain_body)) + 
  geom_histogram(bins = 50, fill = "lightblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Bootstrap distribution regression coefficient brain_body")

```


```{r}
df_boot_mammals %>%
  ggplot(aes(danger)) + 
  geom_histogram(bins = 50, fill = "lightblue", color = "black") +
  theme_minimal() +
  labs(
    title = "Bootstrap distribution regression coefficient for danger")
```


## Compare the confidence intervals

First write code to obtain the bootstrap percentile 95% confidence intervals (see code in the lecture slide) and save the results:

```{r}
CI95_boot <- apply(boot_mammals, 2, function(x) quantile(x, probs = c(.025, .975))) %>% t()
CI95_boot %>% round(3)
```

Display the results of the theoretical 95% confidence interval you obtained earlier. 
```{r}
CI95 %>% round(3)
```

Then compare the bootstrap 95% CI and the theoretical 95% CI. What do the results tell you bout the tenability of the model assumptions? 

```{r}
(CI95_boot  - CI95) %>% round(3)
```

# Advanced

## Make more advanced plots

### Plot the confidence intervals in the bootstrap distribution

In a previous exercise you created histograms of the bootstrap distribution of the regression coefficients for `brain_body` and `danger`. Use the same histograms and add to the plots as vertical lines: 

* the limits for the theoretical and bootstrap 95% confidence intervals. 
* the mean of the bootstrapped regression coefficients.
* the values of the regression coefficients of the fitted model.

With the option `geom_vline()` you can add vertical lines to your plot. 

The package `patchwork` allows you to display two plots in one figure by simply adding a `+` sign between the two plots. 

```{r}
library(patchwork)
df_boot_mammals %>%
  ggplot(aes(brain_body)) + 
  geom_histogram(bins = 60, fill = "lightblue", color = "black") +
  geom_vline(xintercept = fit_mammals$coefficients[[2]], col = "darkblue", linewidth = 1) +
  geom_vline(xintercept = mean(df_boot_mammals$brain_body), col = "orange", linewidth = 1) +
  geom_vline(xintercept = CI95_boot[2,1], col = "orange", linewidth = 1, linetype = "dashed") +
  geom_vline(xintercept = CI95_boot[2,2], col = "orange", linewidth = 1, linetype = "dashed") +
  geom_vline(xintercept = CI95[2,1], col = "darkblue", linewidth = 1, linetype = "dashed") +
  geom_vline(xintercept = CI95[2,2], col = "darkblue", linewidth = 1, linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Distribution bootstrapped regression coefficients brain weight and danger",
    subtitle = "Orange= bootstrap mean and 95%CI , dark blue = theoretical estimate and 95%CI") +
  
df_boot_mammals %>%
  ggplot(aes(danger)) + 
  geom_histogram(bins = 60, fill = "lightblue", color = "black") +
  geom_vline(xintercept = fit_mammals$coefficients[[3]], col = "darkblue", linewidth = 2) +
  geom_vline(xintercept = mean(df_boot_mammals$danger), col = "orange", linewidth = 1) +
  geom_vline(xintercept = CI95_boot[3,1], col = "orange", linewidth = 1, linetype = "dashed") +
  geom_vline(xintercept = CI95_boot[3,2], col = "orange", linewidth = 1, linetype = "dashed") +
  geom_vline(xintercept = CI95[3,1], col = "darkblue", linewidth = 1, linetype = "dashed") +
  geom_vline(xintercept = CI95[3,2], col = "darkblue", linewidth = 1, linetype = "dashed") +
  theme_minimal()
```

## Plot the theoretical and bootstrap distributions

In a 1 x 2 plot array, plot the bootstrap sampling distributions of the regression coefficients two variables `brain_body` and `danger` in red, and the theoretical sampling distribution of the regression coefficients in blue. For the latter, use `stat_function(fun = dnorm, args = list(mean = ..., sd = ...))`, with the arguments `mean` and `sd` derived from the corresponding parameter estimates of the `lm` model. Use the `fit_mammals` and the `sum_fit_mammals` objects to obtain the statistics you need for the plots. 

```{r}

ggplot(df_boot_mammals, aes(x = brain_body)) +
  geom_density(col = "red") +
  stat_function(fun = dnorm, args = list(mean = fit_mammals$coef[2], 
                                         sd = sum_fit_mammals$coefficients[2, 2]),
                col = "blue") +
 
ggplot(df_boot_mammals, aes(x = danger)) +
  geom_density(col = "red") +
  stat_function(fun = dnorm, args = list(mean = fit_mammals$coef[3], 
                                         sd = sum_fit_mammals$coefficients[3, 2]),
                col = "blue") 

```


## Add more functionality to the bootstrap function

Modify the `bootstrap()` function in such a way that instead of the bootstrapped parameters it returns

    -   the bootstrap confidence intervals of the parameters and of the R-squared
    -   a plot of the R-squared sampling distribution

```{r}
bootstrap <- function(formula, data, B = 1000) {  
  
  pars    <- colnames(model.matrix(formula, data = data))
  
  output  <- matrix(nrow = B, ncol = length(pars) + 1,           
                    dimnames = list(NULL, c(pars, "Rsquared")))                 
  
  for (i in 1:B) {
    bootsample  <- slice_sample(.data = data, n = nrow(data), replace = TRUE)
    bootfit     <- lm(formula = formula, data = bootsample)
    output[i, ] <- c(bootfit$coef, summary(bootfit)$r.squared)
  }
  
  return(list(confint = apply(output, 2, function(x) quantile(x, probs = c(.025, .975))),
              gg_rsq  = ggplot(as.data.frame(output), aes(Rsquared)) + geom_density()))
}
```

Apply the `bootstrap` function to a model and data set of your own choice.

```{r}
bootstrap(price ~ carat + cut, 
          mutate(diamonds, cut = factor(cut, ordered = F)) %>% 
            slice_sample(n = 2500))
```



