---
title: "Lab 2b"
subtitle: "Data Visualization"
output: 
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    df_print: paged
    number_sections: false
    theme: paper
    highlight: tango
---


---

For this practical, you will need to load the packages `ggplot2`,
`dplyr`, and `patchwork`. If displaying multiple plot in one figure, use
the R chunk options `fig.width` (default = 7), `fig.height` (default =
5) and/or `fig.asp` (height:width ratio) to display the plots in a nice
format. To avoid messages by `ggplot()` being printed to the html, set
the global chunk option to `message = FALSE`

```{r message=FALSE}
library(ggplot2)
library(dplyr)
library(patchwork)
knitr::opts_chunk$set(message = FALSE)
```

For a quick reference of `ggplot2` functions, consult

-   [ggplot2
    Cheatsheet](https://raw.githubusercontent.com/rstudio/cheatsheets/main/data-visualization.pdf)

-   [ggplot2
    Reference](https://ggplot2.tidyverse.org/reference/index.html)

------------------------------------------------------------------------

# The diamonds data

The dataset for this practical is the `diamonds` dataset from the
package `ggplot2`.

a.  Check the help page of this `tibble` for its content, and display
    the first 10 rows of the data in your HTML.

```{r}

```

The `diamonds` data set has more than 50.000 rows, which is a lot. To
simplify matters, we will work with a random sample of 500 cases.

b.  Set the seed to 100, and randomly sample 500 cases from the dataset
    `diamonds` using the `slice_sample()` function. Save the resulting
    data set as `my_diamonds`.

```{r}

```

# One variable

In the following exercises we display plots numeric and of
categorical/discrete variables. Check the `geoms` overview in the
lecture sheets for the different options.

a.  In a 1 x 3 plot array, plot the distribution of variable `carat`
    using the three different geoms. Add a title with the name of the
    appropriate geom to each plot. The chunk option `fig.asp = .35`
    displays the plots in a nice format.

```{r fig.asp = 0.35}

```

Discrete variables are variables with countable values (i.e. categorical
variables or variables with a small number of integer values). The most
common option for such variables is the bar plot

a.  Display the bar plots of the factors `cut`, `color` and `clarity` in
    a 1 x 3 plot array. Make sure that the labels of the categories are
    readable (i.e., not overlapping).

```{r}

```

# Two variables

The combinations between numeric and categorical/discrete variables
requires different geoms. Check the cheat sheet again for the appropriate
options.

a.  In the first row of one figure, display a box and violin plot of the
    variables `price` and `cut`, an in the second row a density plot.
    Use the `/` symbol to start a new row.

```{r}

```

The variables `color` and `clarity` are both categorical variables, and
their distribution is generally displayed by bar plots.

b.  Display a stacked, a dodged, and a filled bar plot of these
    variables in a 1 x 3 plot array. Use the `plot_layout()` function to
    display the legend and axes labels only once.

```{r}

```

c.  Now display the joint distribution of these two variables with
    `geom_count()`.

```{r}

```

The joint distribution of two continuous variables is best depicted with
a scatter plot.

d.  Display two scatter plots of `price` and `carat` in a 1 x 2 plot
    array, once with different sizes for `price`, and once with
    different shapes for `cut`. Make the points transparent by setting
    `alpha` to an appropriate value.

```{r}

```

e. The plot with different colors for `cut` is not very clear.
Reproduce the plot by adding a `facet_wrap()` layer, and add `price` for
the size of the points.

```{r}

```

f. Display the previous plot with a different theme (choose one that you think looks nice).

```{r}

```

---

End of lab
