---
title: "Lab 3a"
subtitle: "Linear model"
author: ""
output: 
  prettydoc::html_pretty:
    toc: true
    toc_depth: 2
---


For this lab you need to load the packages `dplyr`, `ggplot2` and
`patchwork`.


# Data prepapartion

In this practical, we predict the `price` of diamonds of the `diamonds`
data from the `ggplot2` package. The data has 10 variables and 53940
observations. For that many cases it takes a long time to make a plot.
Therefore we reduce the size of the data a little bit.

a.  Set the seed to 100 and draw a random sample of 500 cases with
    `slice_sample()` function. Save this sample as `my_diamonds`


b.  Because we want to work with dummy variable, we have to convert the
    variables of class `ordered` into variables of class `factor`. Do so
    and save the result.

# Simple regression

We start with a numeric predictor.

a.  Display a scatter plot with a linear regression line for the model
    `price ~ carat`.


b.  Display and interpret the summary of the model and interpret the
    effect of `carat` on `price`.


Now we use a categorical predictor.

c.  Plot the distribution of `price` as a function of the categorical
    variable `cut`.

b.  Display and interpret the summary of the model with `cut` predicting
    `price`.

```{r}
lm(price ~ cut, my_diamonds) %>% summary()
```

# Multiple regression

a.  Display a scatter plot with a linear regression lines for the model
    `price ~ carat + cut`. Does the plot show evidence for interaction?

b.  Display the summary of the interaction model `price ~ carat * cut`,
    and interpret the parameter estimates.

c.  Perform an ANOVA test for the model main effects model and the
    interaction model. Which model fits the data best?

We now add the numeric variable `depth` and the categorical variable
`color` as predictors of `price`.

d.  Perform an ANOVA test for by adding the variables `carat`, `cut`,
    `dpth` and `color` subsequently to the model. What do you conclude?


The `step()` function is specifically designed to find the best model
given a set of predictors.

e.  Apply the `step()` function to the main effects model with all four
    predictors.


f.  Perform a backward step-wise analysis with all main with the predictors `carat`, `color`, `cut`, and `depth`.


g.  Display and interpret the summary this model.

```{r}
summary(lm(price ~ carat + cut + color, my_diamonds))
```

# Advanced

In the previous exercises we did not investigate potential interactions between the predictors.

a.  Use the `step()` function to find the best model when allowing for all pairwise interactions between the predictors. 


b. There is one interaction term in the model. To interpret this interaction, display it a plot.


---

End of lab
