---
title: "Lab 4b: Programming"
subtitle: "Statistical Programming with R"
author: 'Utrecht Summer School'
output: 
  ioslides_presentation:
    smaller: true
    css: custom.css
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment="")
library(tidyverse)
set.seed(123)
```

## Content

1.  Looping

2.  Writing your own function

# Looping

## Loops

Iterates a set of commands multiple times

Loops are used for:

-   simulation studies

-   bootstrapping

-   maximization of non-linear functions

-   and solving many other problems

<br>

Loops save you a lot of typing and prevents making errors

## The `for` loop

The code

```{r eval = F}
for (counter in sequence) {
  
  expressions  # all axpression with curly brackets are executed
  
}
```

-   `counter` variable that subsequently takes the values in the `sequence`

-   `sequence` sequence of value that are used in the `expressions`

-   `expressions` are executed as many times as the length of the `sequence`

## Example

Compute the exponents of the numbers 1 to 10

-   in the 1st loop `i = 1`, in the 2nd `i = 2`, etc.
-   the loop stops after `1 = 10`

```{r}
for (i in 1:10) {
  print(exp(i))
}
```

## Container objects

Object to store the results of each iteration in the loop

-   has to be initialized before the loop starts
-   matrices and lists are best suited for this

```{r}
out <- matrix(nrow = 10, ncol = 2, dimnames = list(1:10, c("x", "exp(x)")))    
for (i in 1:10) {
 out[i, ] <- exp(i)
}
out
```

## Plot example

Display `mpg` on y-axis and all other variables on the x-axis (`fig.asp=.35`)

```{r fig.asp=.35}
par(mfrow = c(2, 5), mar = c(4, 4, 1, 2))
for (i in 2:11) {
  plot(x = mtcars[, i], y = mtcars[, 1])
}
```

## Adding the variable names

```{r fig.asp=.35}
vars <- colnames(mtcars)
par(mfrow = c(2, 5), mar = c(4, 4, 1, 2))
for (i in 2:11) {
  plot(x = mtcars[, i], y = mtcars[, 1], xlab = vars[i], ylab = vars[1])
}
```

## Other loop functions

R functions that perform loops over rows/columns of matrices/data frames

```{r eval= F}
apply(X, MARGIN, FUN) 

sapply(X, FUN)
```

-   `X` is name data frame / matrix

-   `MARGIN` 1 = rows, 2 = columns

-   `FUN` summary function that returns a statistic (e.g. `max`, `mean`, `var`, etc.)

<br>

For some applications easier to use than a loop

## Examples

Get the maximum value of each of the variable in `mtcars`

```{r}
sapply(mtcars, max)
```

Get the `mean` in the rows of `mtcars` (silly example)

```{r}
apply(mtcars, 1, mean)
```

## `sapply` with function arguments

To return the 2.5% and 97.5% percentiles of the variables in `mtcars`

The function quantile returns percentile of vector `x` , e.g.

```{r eval = F}
quantile(x, prob = c(.025, .975))
```

Used in `sapply`

```{r}
sapply(mtcars, quantile, prob = c(.025, .975))
```

# Writing functions

## Write your own function

The function is `function()`

```{r eval=F}
my_function <- function(arguments) {
  
  expressions       
  
  return(output)   
                  
}
```

-   `arguments` are input of the function

-   `expressions` are operations performed on the arguments

-   `output` an object containing the output (e.g. vector, matrix, list, etc.)

-   `return` optional, the object `output` is sufficient

## Tossing a die

A function without arguments that simulates tossing a die

```{r}
die <- function() {
  
  eyes <- sample(1:6, size = 1)
  
  return(eyes)
}
```

```{r}
c(die(), die(), die())
```

## Defining an argument

The argument `n` specifies the number of dice to toss

```{r}
dice <- function(n) {
  
  return(sample(1:6, size = n, replace = TRUE))
}

dice(10)
```

## Multiple returns

A function `dice(n)` returning a list with

-   the outcomes of the `n` throws, their frequencies and their mean

```{r}
dice <- function(n) {
  
  eyes <- sample(1:6, size = n, replace = TRUE) 
  
  return(list(outcomes = eyes,
              freqs    = table(eyes),
              mean     = mean(eyes)))
}
dice(10)
```

## Default arguments

The default is a fair die (each outcome has probability 1/6)

-   the user can change this if so desired

```{r}
dice <- function(n, p = rep(1/6, 6)) {
  
  eyes <- sample(1:6, size = n, replace = TRUE, prob = p)
  
  return(list(outcomes  = eyes,
              frequency = table(eyes),
              mean      = mean(eyes)))
}
dice(100)
```

## Unfair dice

The following command throws 100 unfair dice

-   probabilities 1, 2, 3, 4, 5 is 0.1
-   probability of 6 is 0.5

```{r}
dice(100,  c(rep(.1, 5), .5))
```

# Lab 4a
