---
title: "Operators"
subtitle: "Statistical Programming with R"
author: 'Utrecht Summer School'
output: 
  ioslides_presentation:
    smaller: true
    css: custom.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment="")
set.seed(123)
```

## Content

Data manipulation: 

* Subsetting

* Logical operators


# Data manipulation: subsetting

## Subsetting
In the basic installation of R ("base R") there are three ways to select elements from vectors, matrices, lists and data frames: 

1) **`[ ]`**

2) **`[[ ]]`**

3) **`$`**

## Subsetting with `[ ]`

Square brackets `[ ]` are used to call single elements or more elements.

```{r echo=FALSE}
scores <- c(1/3, 1/5, 2/4, 3/4, 6/7)
scores
```

Select the second element with:
```{r}
scores[2]
```

or the elements 2 to 4 with:
```{r}
scores[2:4]
```


## Subsetting with `[ ]` in matrices

Square brackets `[ ]` are used to call single elements or entire rows and columns.

**`[a, b]`**: **a** refers to the row number(s), **b** refers to the column number(s).

```{r echo=FALSE}
M <- matrix(data=c(1,2,3,4,5,6), nrow=3, ncol=2)
M
```

Get the element in row 2, column 1.

```{r}
M[2, 1]
```

## Subsetting with `[ ]`

Also for data frames:

```{r echo=FALSE}
first_names <- c("Jack", "Anne", "George", "Mike", "Sophie")
transportation <- factor(c("bike", "car", "train", "bus", "walking"))
D <- data.frame(first_names, scores, transportation)
D
```

<div class="columns-2">
Select second row:
```{r}
D[2, ] 
```

Select first column:
```{r}
D[, 1] 
```
</div>


## Subsetting with `[[]]`

The `[[]]` operator selects only one element, for example, in a list:

```{r echo=FALSE}
L <- list(first_names, M, transportation)
L
```

Select first object of the list:
```{r }
L[[1]] 
```

## Subsetting with `$`

Use **`$`** to select elements with name labels in data frames or lists:

```{r echo=FALSE}
D
```


```{r}
D$first_names
```

# Overview subsetting

## Subsetting with `[]`, `$` and `[[]]`

To select elements/rows/columns from an R data object:

| object     | code        | selection                         |
|------------|-------------|-----------------------------------|
| vector     | `v[2:4]`    | elements 2 to 4                   |
| vector     | `v["dog"]`  | element named "dog"               |
| matrix     | `m[, 3]`    | 3rd column                        |
| matrix     | `m[2, 3]`   | element in 2nd row and 3rd column |
| data frame | `df[2, 3]`  | idem                              |
| data frame | `df$dog`    | variable "dog"                    |
| data frame | `df$dog[2]` | 2nd element of variable "dog"     |
| list       | `l[[2]]`    | 2nd list element                  |
| list       | `l$dog`     | list element named "dog"          |

# Logical operators

## Logical operators
Logical operators are signs that evaluate a statement, such as `==`, `<`, `>`, `<=`, `>=`, and `|` (OR) as well as `&` (AND). 

Typing `!` before a logical operator takes the complement of that action. 

Example: which elements of vector `v` are larger than 6?
```{r collapse=T}
(v <- c(1:12))
v > 6
```

## Subsetting with logical operators

Use the logical vector to select elements from a vector
```{r collapse=T}
v
(selection <- v > 6)
```

```{r}
v[selection]
```

Or, more efficient:
```{r}
v[v > 6]
```


## Operators returning `TRUE/FALSE`

| Symbol       | Meaning                                              |
|--------------|------------------------------------------------------|
| `a == b`     | `TRUE` if `a` is equal to `b`, otherwise `FALSE`     |
| `a != b`     | `TRUE` if `a` is not equal to `b`, otherwise `FALSE` |
| `a & b == c` | `TRUE` if `a` and `b` are both equal `c`             |
| `a | b == c` | `TRUE` if `a` or `b` is equal to `c`                 |
| `a > b`      | `TRUE` if `a` is greater than `b`                    |
| `a >= b`     | `TRUE` if `a` is equal to or greater than `b`        |
| `a %in% b`   | `TRUE` if `a` is present *in* `b`                    |




# Lab 1d

## Lab 1d

* Download the files for Lecture and Lab 1d.
* Save the files in your RSummer2024 project folder and maintain a logical structure.

